---
title: "Aggregation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Aggregation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include = FALSE}
library(magrittr)
library(Recca)
```


## Introduction

`Recca` (`R` Energy Conversion Chain Analysis) contains 
many functions useful for aggregating industries and products in energy conversion chains.

The types of aggregation are shown in the following table.

| Function                   | Meaning                                                                                     |
|---------------------------:|:--------------------------------------------------------------------------------------------|
| `primary_aggregates()`     | Aggregates primary energy                                                                   |
| `finaldemand_aggregates()` | Aggregates final demand energy                                                              |
| `region_aggregates()`      | Aggregates regions                                                                          |
| `despecified_aggregates()` | Despecifies industries and products, then aggregates                                        |
| `grouped_aggregates()`     | Groups industries and products, then aggregates                                             |
| `footprint_aggregates()`   | Isolates each final demand product and sector, then aggregates by primary and final demand  |
| `effects_aggregates()`     | Isolates each product in the resources matrix, then aggregates by primary and final demand  |

This vignette describes those functions and demonstrates their use.


## `primary_aggregates()`

`primary_aggregates()` creates a column of primary energy sums coming from the industries 
specified in the `p_industries` argument.
These industries are often found in the **R**, **V**, and **Y** matrices 
of the PSUT framework.
Typical `p_industries` are Resources, Imports, and Exports.
In the example below, 
primary aggregates are calculated 
for the four rows of the `UKEnergy2000mats` data frame.

In the first example,
only 1 aggregate column is created ("EX.p"), because there is no
difference between net and gross aggregation for primary energy.
Net and gross primary aggregates are identical.

```{r}
library(matsbyname)
p_industries <- c("Resources - Crude", "Resources - NG")
# Calculate primary total aggregates
ECCs <- UKEnergy2000mats %>%
  tidyr::pivot_wider(names_from = "matrix.name", values_from = "matrix")
ECCs
res <- ECCs %>%
  Recca::primary_aggregates(p_industries = p_industries, by = "Total")
tibble::as_tibble(res)
res[[Recca::aggregate_cols$aggregate_primary]]
```

In the next example,
both net and gross aggregation columns are created
("EX.p_net" and "EX.p_gross"),
for compatibility with the `finaldemand_aggregates()` function
discussed below.

```{r}
res2 <- ECCs %>%
  Recca::primary_aggregates(p_industries = p_industries,
                            add_net_gross_cols = TRUE,
                            by = "Total")
tibble::as_tibble(res2)
res2[[Recca::aggregate_cols$net_aggregate_primary]]
res2[[Recca::aggregate_cols$gross_aggregate_primary]]
```


## `finaldemand_aggregates()`

`finaldemand_aggregates()` is similar to `primary_aggregates()`, 
except that it aggregates energy at the final demand stage in the **Y** matrix.

```{r}
fd_sectors <- c("Residential", "Transport")
res <- ECCs %>%
  Recca::finaldemand_aggregates(fd_sectors = fd_sectors, by = "Sector")
tibble::as_tibble(res)
res[[Recca::aggregate_cols$net_aggregate_demand]]
res[[Recca::aggregate_cols$gross_aggregate_demand]]
```


## `region_aggregates()`



## `despecified_aggregates()`



## `grouped_aggregates()`



## `footprint_aggregates()`



## `effects_aggregages()`





## Conclusion





