---
title: "Aggregation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Aggregation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include = FALSE}
library(magrittr)
library(Recca)
```


## Introduction

`Recca` (`R` Energy Conversion Chain Analysis) contains 
many functions useful for aggregating industries and products in energy conversion chains.

The types of aggregation are shown in the following table.

| Function                   | Meaning                                                                                                  |
|---------------------------:|:---------------------------------------------------------------------------------------------------------|
| `primary_aggregates()`     | Aggregates primary energy                                                                                |
| `finaldemand_aggregates()` | Aggregates final demand energy, the last stage in an energy conversion chain (ECC)                       |
| `region_aggregates()`      | Aggregates regions                                                                                       |
| `despecified_aggregates()` | Despecifies industries and products, then aggregates                                                     |
| `grouped_aggregates()`     | Groups industries and products, then aggregates                                                          |
| `pfu_aggregates()`         | Aggregates to primary, final, and useful stages of an ECC                                                |

This vignette describes those functions and demonstrates their use.


## `primary_aggregates()`

`primary_aggregates()` creates a column of primary energy sums coming from the industries 
specified in the `p_industries` (primary industries) argument.
These primary industries are often found in the **R**, **V**, and **Y** matrices 
of the PSUT framework.
Typical `p_industries` are Resources, Imports, and Exports.
In the examples below, 
primary aggregates are calculated 
for the four rows of the `UKEnergy2000mats` data frame.

In the first example,
only one aggregate column is created ("EX.p"), because there is no
difference between net and gross aggregation for primary energy.
Net and gross primary aggregates are identical.

```{r}
library(matsbyname)
p_industries <- c("Resources [of Crude]", "Resources [of NG]")
# Calculate primary total aggregates
ECCs <- UKEnergy2000mats %>%
  tidyr::pivot_wider(names_from = "matrix.name", values_from = "matrix")
ECCs
res <- ECCs %>%
  Recca::primary_aggregates(p_industries = p_industries, by = "Total")
tibble::as_tibble(res)
res[[Recca::aggregate_cols$aggregate_primary]]
```

In the next example,
both net and gross aggregation columns are created
("EX.p_net" and "EX.p_gross"),
for compatibility with the `finaldemand_aggregates()` function
discussed below.

```{r}
res2 <- ECCs %>%
  Recca::primary_aggregates(p_industries = p_industries,
                            add_net_gross_cols = TRUE,
                            by = "Total")
tibble::as_tibble(res2)
res2[[Recca::aggregate_cols$net_aggregate_primary]]
res2[[Recca::aggregate_cols$gross_aggregate_primary]]
```


## `finaldemand_aggregates()`

`finaldemand_aggregates()` is similar to `primary_aggregates()`, 
except that it aggregates energy at the final demand stage in the **Y** matrix.
Both net and gross aggregate final demand are calculated.
Gross final demand includes energy industry own use (EIOU).
Net final demand does not.

```{r}
fd_sectors <- c("Residential", "Transport")
res <- ECCs %>%
  Recca::finaldemand_aggregates(fd_sectors = fd_sectors, by = "Sector")
tibble::as_tibble(res)
res[[Recca::aggregate_cols$net_aggregate_demand]]
res[[Recca::aggregate_cols$gross_aggregate_demand]]
```


## `region_aggregates()`

`region_aggregates()` sums regions according to `many_colname` and `few_colname`.
To demonstrate this function, we need to modify the The example data frame 
slightly. 
It contains only one country ("GBR"), but it should contain
different countries and a "few" column.
Furthermore, the last stage and energy type columns should be the same
where aggregation is required.
The following code produces this modification to pretend that 
rows of the data frame apply to different countries.

```{r}
ECCs_for_region_agg <- ECCs |> 
  dplyr::mutate(
    Country = c("USA", "GBR", "CAN", "FRA"), 
    Continent = c("NoAmr", "Europe", "NoAmr", "Europe"), 
    Energy.type = "E", 
    Last.stage = c("Final", "Services", "Final", "Services")
  )
ECCs_for_region_agg |> 
  dplyr::select(Country, Continent, Energy.type, Last.stage)
```

Given this modification, 
we aggregate to continents with the following code.

```{r}
continent_aggregations <- ECCs_for_region_agg |> 
  Recca::region_aggregates(many_colname = "Country", few_colname = "Continent")
continent_aggregations
```

By default, `many_colname` is replaced by the common values in `few_colname`.
A simple rename can override this default behaviour.

```{r}
continent_aggregations |> 
  dplyr::rename(
    Continent = Country
  )
```


## `despecified_aggregates()`

When the row and column names of the PSUT matrices
are "specified," they will look like "Resources [of Crude]"
or "NG [from Wells]".
Many labels can have a "noun [preposition object]" structure,
which we call "specified."
`despecified_aggregates()` 
eliminates the specification in the row and/or column names, 
keeping the desired `piece` (by default the "noun"),
and aggregates (sums) rows or columns 
with resulting identical names.
See the comments in the examples below.

```{r}
despecified_aggs <- ECCs |> 
  # Accept the default "noun" aggregation
  # and the default "_aggregated" suffix for 
  # aggregated columns.
  despecified_aggregates()

# This is the original R matrix.
# Its row names are specified.
ECCs$R[[1]]

# This is the despecified R matrix.
# Note the single Resources row.
despecified_aggs$R_aggregated[[1]]

# Here is an original use matrix.
# Many rows are specified.
ECCs$U[[2]]

# Here is the corresponding despecified use matrix. 
# None of the rows are specified.
despecified_aggs$U_aggregated[[2]]

# Here is an original make matrix
# with many specified columns.
ECCs$V[[3]]

# None of the columns are specified
# in the despecified version.
despecified_aggs$V_aggregated[[3]]

# This original final demand matrix
# has specified rownames
# that provide units.
ECCs$Y[[4]]

# Despecifying this matrix eliminates the units,
# resulting in information loss.
despecified_aggs$Y_aggregated[[4]]
```

As the last example (with the final demand matrix, **Y**) shows, 
aggregating despecified row and column names
can lead to information loss.
Thus, `despecified_aggregates()`
will normally be called only at the last step
in a calculation chain.


## `grouped_aggregates()`

Grouping provides the capability to aggregate 
specific energy products to classes of energy
and groupings of processing stages or final demand categories.
For example, 
Anthracite and Brown coal can be grouped to 
Coal and coal products. 
An aggregation map is required.

To demonstrate grouped aggregation,
we first establish an aggregation map.
An aggregation map is 
a named list where list members
are aggregated to member names. 

```{r}
agg_map <- list(`Crude oil` = c("Crude", "Crude [from Dist.]", "Crude [from Fields]"), 
                `Oil and oil products` = c("Diesel", "Diesel [from Dist.]", 
                                           "Petrol", "Petrol [from Dist.]"), 
                NG = c("NG", "NG [from Dist.]", "NG [from Wells]"),
                Electricity = c("Elect", "Elect [from Grid]"))
```

The aggregation map can be used with `grouped_aggregates()` 
to do the desired aggregation.
Aggregation columns are added 
to the right of the `.sut_data` data frame.

```{r}
# Here is an original use matrix.
ECCs$U[[1]]

# Aggregate to the desired groups.
ECCs_grouped_aggs <- ECCs |> 
  grouped_aggregates(aggregation_map = agg_map)

# Here is the aggregated use matrix.
# Note that the rows are summed
# and named according to the agg_map.
ECCs_grouped_aggs$U_aggregated[[1]]
```


## `pfu_aggregates()`

Aggregating to primary, final, and useful stages
of the energy conversion chain (ECC) can be accomplished
with `pfu_aggregates()`.
Aggreagtes are added in columns to the right of the `.sutdata` data frame.
Primary industries and final demand sectors
are specified as character vectors in the 
`p_industries` and `fd_sectors` arguments.

```{r}
# Get the defaul separator for column names
sep <- Recca::all_stages$last_stage_sep
# Set primary industry names and final demand sector names
p_industries <- c("Resources [of Crude]", "Resources [of NG]")
fd_sectors <- c("Residential", "Transport", "Oil fields")
# Calculate TOTAL aggregates
pfu_aggs_total <- UKEnergy2000mats |>
  tidyr::pivot_wider(names_from = matrix.name, values_from = matrix) |>
  pfu_aggregates(p_industries = p_industries, fd_sectors = fd_sectors,
                 by = "Total")
# Look at the column names.
# The naming scheme is 
# EX.stage.gn___lsStage
# EX = energy or exergy aggregate (depending on the value in the Energy.type column)
# stage = the stage at which the aggregation is computed
# gn = whether the aggregate is "gross" or "net"
# ____ls = the separator (Recca::all_stages$last_stage_sep)
# Stage = The last stage in the ECC for this column.
names(pfu_aggs_total)
# Check some aggregation values.
# Note that not all aggregations are available, 
# based on the structure of the ECC.

# Net primary energy aggregation when last stage is Final
pfu_aggs_total |>
  dplyr::filter(Energy.type == "E") |>
  magrittr::extract2("EX.p_net___lsFinal") |>
  magrittr::extract2(1)
# Net primary exergy aggregation when last stage is Services
pfu_aggs_total |>
  dplyr::filter(Energy.type == "X") |>
  magrittr::extract2("EX.p_net___lsServices") |>
  magrittr::extract2(1)
# Gross final energy aggregation when last stage is Useful
pfu_aggs_total |>
  dplyr::filter(Energy.type == "E") |>
  magrittr::extract2("EX.f_gross___lsFinal") |>
  magrittr::extract2(1)
# Gross useful energy aggregation when last stage is Useful
pfu_aggs_total |>
  dplyr::filter(Energy.type == "E") |>
  magrittr::extract2("EX.u_gross___lsUseful") |>
  magrittr::extract2(1)
# Net useful energy aggregation when last stage is Useful
pfu_aggs_total |>
  dplyr::filter(Energy.type == "E") |>
  magrittr::extract2("EX.u_net___lsUseful") |>
  magrittr::extract2(1)
# Calculate PRODUCT aggregates
pfu_aggs_product <- UKEnergy2000mats |>
  tidyr::pivot_wider(names_from = matrix.name, values_from = matrix) |>
  pfu_aggregates(p_industries = p_industries, fd_sectors = fd_sectors,
                 by = "Product")
# Net primary energy aggregation when last stage is Final
# Note that all results are now vectors which 
# show aggregation by product
pfu_aggs_product |>
  dplyr::filter(Energy.type == "E") |>
  magrittr::extract2("EX.p_net___lsFinal") |>
  magrittr::extract2(1)
pfu_aggs_product |>
  dplyr::filter(Energy.type == "E") |>
  magrittr::extract2("EX.u_net___lsUseful") |>
  magrittr::extract2(1)
# Calculate INDUSTRY aggregates
# Note that all results are now vectors which 
# show aggregation by industry
pfu_aggs_industry <- UKEnergy2000mats |>
  tidyr::pivot_wider(names_from = matrix.name, values_from = matrix) |>
  pfu_aggregates(p_industries = p_industries, fd_sectors = fd_sectors,
                 by = "Industry")
pfu_aggs_industry |>
  dplyr::filter(Energy.type == "E") |>
  magrittr::extract2("EX.p_net___lsFinal") |>
  magrittr::extract2(1)
pfu_aggs_industry |>
  dplyr::filter(Energy.type == "E") |>
  magrittr::extract2("EX.u_net___lsUseful") |>
  magrittr::extract2(1)
```


## Conclusion

Several functions in the `Recca` package assist with aggregations
in energy conversion chains.
`primary_aggregates()`,
`finaldemand_aggregates()`,
`region_aggregates()`,
`despecified_aggregates()`,
`grouped_aggregates()`, and
`pfu_aggregates()`
all assist with different types of aggregations.
Please see help for these functions for more details.
