---
title: "Aggregation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Aggregation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include = FALSE}
library(magrittr)
library(Recca)
```


## Introduction

`Recca` (`R` Energy Conversion Chain Analysis) contains 
many functions useful for aggregating industries and products in energy conversion chains.

The types of aggregation are shown in the following table.

| Function                   | Meaning                                                                                     |
|---------------------------:|:--------------------------------------------------------------------------------------------|
| `primary_aggregates()`     | Aggregates primary energy                                                                   |
| `finaldemand_aggregates()` | Aggregates final demand energy, the last stage in an energy conversion chain (ECC)          |
| `region_aggregates()`      | Aggregates regions                                                                          |
| `despecified_aggregates()` | Despecifies industries and products, then aggregates                                        |
| `grouped_aggregates()`     | Groups industries and products, then aggregates                                             |
| `footprint_aggregates()`   | Isolates each final demand product and sector, then aggregates by primary and final demand  |
| `effects_aggregates()`     | Isolates each product in the resources matrix, then aggregates by primary and final demand  |

This vignette describes those functions and demonstrates their use.


## `primary_aggregates()`

`primary_aggregates()` creates a column of primary energy sums coming from the industries 
specified in the `p_industries` (primary industries) argument.
These primary industries are often found in the **R**, **V**, and **Y** matrices 
of the PSUT framework.
Typical `p_industries` are Resources, Imports, and Exports.
In the examples below, 
primary aggregates are calculated 
for the four rows of the `UKEnergy2000mats` data frame.

In the first example,
only 1 aggregate column is created ("EX.p"), because there is no
difference between net and gross aggregation for primary energy.
Net and gross primary aggregates are identical.

```{r}
library(matsbyname)
p_industries <- c("Resources [of Crude]", "Resources [of NG]")
# Calculate primary total aggregates
ECCs <- UKEnergy2000mats %>%
  tidyr::pivot_wider(names_from = "matrix.name", values_from = "matrix")
ECCs
res <- ECCs %>%
  Recca::primary_aggregates(p_industries = p_industries, by = "Total")
tibble::as_tibble(res)
res[[Recca::aggregate_cols$aggregate_primary]]
```

In the next example,
both net and gross aggregation columns are created
("EX.p_net" and "EX.p_gross"),
for compatibility with the `finaldemand_aggregates()` function
discussed below.

```{r}
res2 <- ECCs %>%
  Recca::primary_aggregates(p_industries = p_industries,
                            add_net_gross_cols = TRUE,
                            by = "Total")
tibble::as_tibble(res2)
res2[[Recca::aggregate_cols$net_aggregate_primary]]
res2[[Recca::aggregate_cols$gross_aggregate_primary]]
```


## `finaldemand_aggregates()`

`finaldemand_aggregates()` is similar to `primary_aggregates()`, 
except that it aggregates energy at the final demand stage in the **Y** matrix.
Both net and gross aggregate final demand are calculated.
Gross final demand includes energy industry own use (EIOU).
Net final demand does not.

```{r}
fd_sectors <- c("Residential", "Transport")
res <- ECCs %>%
  Recca::finaldemand_aggregates(fd_sectors = fd_sectors, by = "Sector")
tibble::as_tibble(res)
res[[Recca::aggregate_cols$net_aggregate_demand]]
res[[Recca::aggregate_cols$gross_aggregate_demand]]
```


## `region_aggregates()`

`region_aggregates()` sums regions according to `many_colname` and `few_colname`.
To demonstrate this function, we need to modify the The example data frame 
slightly. 
It contains only one country ("GBR"), but it should contain
different countries and a "few" column.
Furthermore, the last stage and energy type columns should be the same
where aggregation is required.
The following code produces this modification to pretend that 
rows of the data frame apply to different countries.

```{r}
ECCs_for_region_agg <- ECCs |> 
  dplyr::mutate(
    Country = c("USA", "GBR", "CAN", "FRA"), 
    Continent = c("NoAmr", "Europe", "NoAmr", "Europe"), 
    Energy.type = "E", 
    Last.stage = c("Final", "Services", "Final", "Services")
  )
ECCs_for_region_agg |> 
  dplyr::select(Country, Continent, Energy.type, Last.stage)
```

Given this modification, 
we aggregate to continents with the following code.

```{r}
continent_aggregations <- ECCs_for_region_agg |> 
  Recca::region_aggregates(many_colname = "Country", few_colname = "Continent")
continent_aggregations
```

By default, `many_colname` is replaced by the common values in `few_colname`.
A simple rename can override this default behaviour.

```{r}
continent_aggregations |> 
  dplyr::rename(
    Continent = Country
  )
```



## `despecified_aggregates()`



## `grouped_aggregates()`



## `footprint_aggregates()`



## `effects_aggregages()`


## `pfu_aggregates()`


## Conclusion





