---
title: "Aggregation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Aggregation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include = FALSE}
library(magrittr)
library(Recca)
```


## Introduction

`Recca` (`R` Energy Conversion Chain Analysis) contains 
many functions useful for aggregating industries and products in energy conversion chains.

The types of aggregation are shown in the following table.

| Function                   | Meaning                                                                                     |
|---------------------------:|:--------------------------------------------------------------------------------------------|
| `primary_aggregates()`     | Aggregates primary energy                                                                   |
| `finaldemand_aggregates()` | Aggregates final demand energy                                                              |
| `region_aggregates()`      | Aggregates regions                                                                          |
| `despecified_aggregates()` | Despecifies industries and products, then aggregates                                        |
| `grouped_aggregates()`     | Groups industries and products, then aggregates                                             |
| `footprint_aggregates()`   | Isolates each final demand product and sector, then aggregates by primary and final demand  |
| `effects_aggregates()`     | Isolates each product in the resources matrix, then aggregates by primary and final demand  |

This vignette describes those functions and demonstrates their use.

