---
title: "Balances"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Balances}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE}
library(dplyr)
library(IEATools)
library(Recca)
```


## Introduction

When working with material and energy conversion chains 
in the `Recca` package with `matsindf` data frames,
it is helpful to calculate several types of balances.
This vignette describes those balances and shows
when you might want to use each of them.
Balances include

* inter-industry balances,
* intra-industry balances, and
* the $\mathbf{W}$ matrix.


### Terminology

`Recca` assists with matrix-based input-output calculations
when the data are organized into the 
$\mathbf{RUVY}$ matrices of the 
physical supply-use table (PSUT) framework.
A generic processing stage is called an "industry."
The name for items that flow between industries is "product." 
A "balance" is any calculation that compares inputs to outputs.
Balances can be computed for all types of products, 
in particular materials (quantified in mass or exergy units)
and energy (quantified as energy or exergy).

In the PSUT framework, 
the $\mathbf{R}$ matrix is an industry-by-product resource matrix.
The $\mathbf{U}$ matrix is an product-by-industry use matrix.
The $\mathbf{V}$ matrix is an industry-by-product make matrix.
The $\mathbf{Y}$ matrix is an product-by-industry final demand matrix.

In this vignette, matrices are represented by boldface 
capital letters, such as $\mathbf{R}$.
Vectors are assumed to be column vectors and
represented by boldface lowercase letters, 
such as the identity vector $\mathbf{i}$.


### Data

In the examples that follow, 
we use the energy conversion chain in `UKEnergy2000mats`.

```{r}
ecc <- UKEnergy2000mats |> 
  tidyr::pivot_wider(names_from = matrix.name, 
                     values_from = matrix) |> 
  dplyr::filter(.data[[IEATools::iea_cols$last_stage]] %in%
                  c(IEATools::last_stages$final, IEATools::last_stages$useful))
dplyr::glimpse(ecc)
```


## Balances

The following subsections explain
each type of balance and how to calculate
it with the `Recca` package.


### Inter-industry balances check for product consistency

In any conversion chain, 
products that leave one industry must enter another industry.
These between industry (or inter-industry) balances
should be verified for every conversion chain.

Inter-industry balances are obtained
by the following computation.

$$[(\mathbf{R} + \mathbf{V})^\mathrm{T} 
     - (\mathbf{U} + \mathbf{Y})] \, \mathbf{i}$$

If the conversion chain is consistent, 
the result should be the $\mathbf{0}$ vector 
with products in rows.

To compute inter-industry balances within the PSUT framework
using the `Recca` package, 
use the following code.

```{r}
inter_industry_balances <- ecc |> 
  calc_inter_industry_balance()
```

The result is added as a new column,
named "SUTInterIndustryBalance" by default.

```{r}
glimpse(inter_industry_balances)
```

The vectors in SUTInterIndustryBalance have products in rows.

```{r}
inter_industry_balances$SUTInterIndustryBalance[[1]]
inter_industry_balances$SUTInterIndustryBalance[[2]]
```

As expected, these vectors are the $\mathbf{0}$ vector to within 
reasonable precision.

If you want to check the inter-industry balances for consistency, 
use `verify_inter_industry_balance()`.

```{r}
inter_industry_balances |> 
  verify_inter_industry_balance() |> 
  glimpse()
```

If a product is out of balance 
by more than `tol` (by default `1e-6`), 
a warning occurs and 
`FALSE` appears in the new column, 
"SUTInterIndustryBalanced."


### Intra-industry balances calculate losses

A second type of balance
(the intra-industry balance)
computes differences _across_ industries.
Typically, the intra-industry balance computes 
inputs less outputs, and 
the result is losses (wastes) as a positive number.

Intra-industry balances
are obtained by the following computation.

$$(\mathbf{U}^\mathrm{T} - \mathbf{V}) \, \mathbf{i}$$

The result is a column vector with industries in rows.

The meaning of intra-industry balances
depends upon the construction
of the matrices.
Key factors are whether or not the matrices include losses
and the type of data in the matrices.
The following table shows meanings.
$\mathbf{0}$ indicates the result is the zero vector;
$\mathbf{m}$ indicates the result is a mass vector;
$\mathbf{e}$ indicates the result is an energy vector;
$\mathbf{b}$ indicates the result is a mass exergy vector;
$\mathbf{x}$ indicates the result is an energy exergy vector; and
$\mathbf{bx}$ indicates the result is a the sum of mass and energy exergy vectors.
Subscript $L$ indicates losses (or wastes), and 
subscript $D$ indicates exergy destroyed by the industry.
MCC indicates a material conversion chain.
ECC indicates an energy conversion chain.
BCC indicates a material exergy conversion chain.
XCC indicates an energy exergy conversion chain.
BXCC indicates a combined material and energy exergy conversion chain.


| Matrices contain              | Matrices include losses (wastes)  | Matrices exclude losses (wastes) |
| :---------------------------: | :-------------------------------: | :------------------------------: |
| Mass (MCC)                    | $\mathbf{0}$                      | $\mathbf{m}_L$                   |
| Energy (ECC)                  | $\mathbf{0}$                      | $\mathbf{e}_L$                   |
| Mass exergy (BCC)             | $\mathbf{b}_D$                    | $\mathbf{b}_D + \mathbf{b}_L$    |
| Energy exergy (XCC)           | $\mathbf{x}_D$                    | $\mathbf{x}_D + \mathbf{x}_L$    |
| Mass and energy exergy (BXCC) | $\mathbf{bx}_D$                   | $\mathbf{bx}_D + \mathbf{bx}_L$  |



### The value added matrix ($\mathbf{W}$)




### Deprecated functions

#### `verify_SUT_energy_balance()`






## Conclusiion
































