---
title: "new_* functions"
author: "Matthew Kuperus Heun"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: recca.bib
vignette: >
  %\VignetteIndexEntry{new_* functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(dplyr)
library(Recca)
library(magrittr)
library(matsbyname)
library(matsindf)
library(tidyr)
```

<!-- Establish some helpful LaTeX shortcuts for equations -->
\newcommand{\transpose}[1]{#1^\mathrm{T}}
\newcommand{\inverse}[1]{#1^{\mathrm{-}1}}
\newcommand{\mat}[1]{\mathbf{#1}}
\newcommand{\colvec}[1]{\mathbf{#1}}
\newcommand{\rowvec}[1]{\transpose{\colvec{#1}}}
\newcommand{\inversetranspose}[1]{\transpose{\left( \inverse{\mat{#1}} \right)}}
\newcommand{\transposeinverse}[1]{\inverse{\left( \transpose{\mat{#1}} \right)}}
\newcommand{\hatinv}[1]{\inverse{\widehat{#1}}}

## Introduction

`Recca` is an `R` package that builds upon input-output (I-O) analysis to analyze energy conversion chains (ECCs).
Historically, input-output (I-O) analysis has been used to analyze "what-if" scenarios involving
changes to the production structure of an economy. 
A common scenario asks "how will will the structure of intermediate industries
(represented by $\mat{U}$ and $\mat{V}$ matrices) change when final demand (represented by the $\mat{Y}$ matrix) changes?"
In the economic arena, Leontief's early work answered this question 
via a matrix inversion that now bears his name,
the "Leontief inverse ($\mat{L}$)" [@Leontief:1941].
Analysts with an I-O background will know the famous equation $\mat{L} = (\mat{I} - \mat{A})^{-1}$.

The `Recca` package provides functions that perform several "what if" analyses for the analyst.
Each function has the name `new_*`, where `*` indicates the changed matrix or portion of the ECC.
Some functions have the suffix `_ps` to indicate the functions assume perfect substitution (`ps`)
of industry inputs throughout the ECC.
The `new_*` functions return the resource ($\mat{R}$), use ($\mat{U}$), make ($\mat{V}$), and final demand ($\mat{Y}$) matrices
that would be present under the conditions given by arguments to the `new_*` functions.
Arguments for "new" versions of vectors or matrices are given the suffix `_prime` in the `new_*` functions.
For example, the updated version of the use ($\mat{U}$) matrix is called `U_prime`.

For the examples that follow, we'll use one ECC from the `UKEnergy2000mats` data frame
which represents primary and final energy flows through a portion of the UK in the year 2000.

```{r}
library(dplyr)
library(magrittr)
library(matsbyname)
library(matsindf)
library(Recca)
library(tidyr)
OurECC <- UKEnergy2000mats %>% 
  filter(Last.stage == "final") %>% 
  spread(key = "matrix.name", value = "matrix")
glimpse(OurECC)
```

After a short discussion of the inputs to the `new_*` functions, 
each type of analysis is described in turn.


## Inputs to `new_*` functions

Inputs to the `new_*` functions include
matrices that describe the ECC and 
matrices that describe the input-output structure of the ECC.


### Matrices that describe the ECC

The `new_*` functions take as inputs 
resource ($\mat{R}$), 
use ($\mat{U}$), 
make ($\mat{V}$), and 
final demand ($\mat{Y}$) matrices
that describe the ECC.
Earlier versions of `Recca` included the resource matrix ($\mat{R}$) 
with the make ($\mat{V}$) matrix.
To separate the resource matrix $\mat{R}$ from the make matrix ($\mat{V}$),
use the `separate_RV()` function.

```{r}
UKEnergyMatsWithR <- OurECC %>% 
  # The following code is necessary until R matrices are included
  # in the UKEnergy2000mats data frame.
  rename(
    R_plus_V = V
  ) %>%
  separate_RV()
glimpse(UKEnergyMatsWithR)
```

To look at the ECC, we can draw a Sankey diagram. 
A Sankey diagram for the first row of the `UKEnergyMatsWithR` data frame is given below.

```{r, fig.width = 7, fig.height = 4}
make_sankey(UKEnergyMatsWithR) %>% 
  extract2("Sankey") %>% # Extracts the "Sankey" column
  extract2(1) # Extracts the first row of the "Sankey" column
```


### Matrices that describe the input-output structure of the ECC

Each `new_*` function requires basic knowledge about the I-O structure of the ECCs
in the form of named vectors and matrices.
The `calc_io_mats` function is the simplest way to obtain the I-O structure
which appears as additional columns appended at the right of the data frame.
(The names of the additional columns are given by 
arguments to the `calc_io_mats()` function.)

The first step in any "what-if" analysis is to calculate the input-output 
structure of the energy conversion chain.
The `calc_io_mats()` function adds several columns to the `UKEnergyMatsWithR` data frame.

```{r}
IO_mats <- UKEnergyMatsWithR %>% 
  # Obtain the I-O structure of the ECCs
  calc_io_mats()
# Note additional columns appended to the data frame.
glimpse(IO_mats)
```

With the I-O structure of the ECC determined
(and represented by the $\colvec{q}$, $\colvec{f}$, and $\colvec{g}$ vectors and the 
$\mat{W}$, $\mat{Z}$, $\mat{K}$, $\mat{C}$, $\mat{D}$, $\mat{A}$, 
$\mat{L_{pxp}}$, and $\mat{L_{ixp}}$ matrices), 
we can use the `new_*` functions to ask "what if" questions.


## What if final demand changes? Function new_Y()

The classic I-O question is "what if final demand ($\mat{Y}$) changes?"
`Recca` has a function for that, `new_Y()`.
The `new_Y()` function looks upstream in the ECC to determine the effects of a change in fina demand.
To demonstrate, we'll double the `Residential` final demand 
for electricity from the grid (`Elect - Grid`) in our ECC.

```{r}
double <- function(x){
  2*x
}

DoubleRes <- IO_mats %>% 
  mutate(
    Y_prime = elementapply_byname(double, a = Y, row = "Elect - Grid", col = "Residential")
  )
DoubleRes$Y[[1]]
DoubleRes$Y_prime[[1]]
```

Now we can use the `new_Y()` function to estimate the ECC with the larger final demand for residential electricity.
The original Sankey diagram is:

```{r, fig.width = 7, fig.height = 4}
make_sankey(UKEnergyMatsWithR) %>% 
  extract2("Sankey") %>% # Extracts the "Sankey" column
  extract2(1) # Extracts the first row of the "Sankey" column
```

The new Sankey diagram is:

```{r, fig.width = 7, fig.height = 4}
DoubleRes <- DoubleRes %>% 
  new_Y()
make_sankey(DoubleRes, U = "U_prime", V = "V_prime", Y = "Y_prime") %>% 
  extract2("Sankey") %>% 
  extract2(1)
```

Output from the electric grid is clearly larger now than before.
Also, upstream demand for natural gas is now larger than the upstream demand for crude,
whereas in the original Sankey diagram, demand for crude was larger than the demand for natural gas.


## What if resources change? Function new_R_ps()

`newY()` estimates the upstream effects of a change to final demand.
But what if we want to understand the downstream effects of a change in resource availability?
The function `new_R_ps()` supplies the answer. 
`new_R_ps()` takes as input the input-output structure of an ECC 
and the efficiency of every intermediate industry.
After discussing the resource matrix and resoure vectors, 


### Resource matrices and vectors

The rate of resource extraction is given by the $\mat{R}$ matrix or
by the $\colvec{r}$ vector.
The resource matrix ($\mat{R}$) is Industry$\times$Product.
Its industries are usually named `Resources - <Product>`,
where `<Product>` is the name of the Product produced by that industry.
(`clean_byname()` removes `0` rows and columns for clarity.)

```{r}
UKEnergyMatsWithR$R[[1]] %>% 
  clean_byname()
```

The resource vector ($\colvec{r}$) is a Product column vector
containing Products that are resources.

```{r}
rvecs <- UKEnergyMatsWithR %>% 
  transmute(
    r = clean_byname(R) %>% 
      colsums_byname() %>% 
      transpose_byname()
  )
rvecs$r[[1]]
```


### Industry effiencies and a new resource matrix

The baseline $\mat{R}$ matrix is shown above.
We can calculate the efficiency of every Industry in the ECC
using the `calc_eta_i()` function.
Let's say we want to estimate the effect of 
a 50 % reduction in the production rate of crude oil. 
So, we cut `Resources - Crude` in half.
To execute this reduction, we'll use the `matsbyname::elementapply_byname()` function.

```{r}
half <- function(x){
  x/2
}
HalfCrude <- IO_mats %>% 
  calc_eta_i() %>% 
  mutate(
    R_prime = elementapply_byname(half, a = R, row = "Resources - Crude", col = "Crude")
  )
HalfCrude$R_prime[[1]] %>% clean_byname()
```

Now we use the `new_R()` function to estimate changes in the ECC.
But we need to calculate the efficiency of all sectors in the ECC first
using the `calc_eta_i()` function.

```{r}
HalfCrude <- HalfCrude %>% 
  new_R_ps()
glimpse(HalfCrude)
```

After re-calculating the `HalfCrude` ECC
(specifically, after calculating `U_prime`, `V_prime`, and `Y_prime`),
we can visualize the effect of lower crude oil availability
with a modified Sankey diagram.

First, we have the original Sankey diagram.

```{r, fig.width = 7, fig.height = 4}
make_sankey(UKEnergyMatsWithR) %>% 
  extract2("Sankey") %>% # Extracts the "Sankey" column
  extract2(1) # Extracts the first row of the "Sankey" column
```

And here is the modified Sankey diagram showing the much smaller `Resources - Crude` extraction rate.

```{r, fig.width = 7, fig.height = 4}
make_sankey(HalfCrude, R = "R_prime", U = "U_prime", V = "V_prime", Y = "Y_prime") %>% 
  extract2("Sankey") %>% # Extracts the "Sankey" column
  extract2(1) # Extracts the first row of the "Sankey" column
```

Note that both ECCs are in complete energy balance.

```{r}
UKEnergyMatsWithR %>% 
  verify_SUT_energy_balance_with_units(S_units = "S_units", U = "U", V = "R_plus_V", Y = "Y") %>% 
  select(Country, Year, Energy.type, Last.stage, .SUT_prod_energy_balance, .SUT_ind_energy_balance)
```


```{r}
HalfCrude %>% 
  filter(!is.na(K)) %>% 
  mutate(
    R_plus_V_prime = sum_byname(R_prime, V_prime)
  ) %>% 
  verify_SUT_energy_balance(U = "U_prime", V = "R_plus_V_prime", Y = "Y_prime") %>% 
  select(Country, Year, Energy.type, Last.stage, .SUT_energy_balance)
```

The Sankey diagram clearly shows that 
reducing the availability of crude oil (`Resources - crude`)
means that less `Transport` final demand can be satisfied.


## What if perfectly-substitutable inputs to an intermediate industry change? Function new_k_ps()

`new_Y()` estimates the upstream effects of a change in final demand.
`new_R_ps()` estimates the downstream effects of a change in resource availability.
But what if we want to estimate the effects of a change to the inputs of an intermediate industry?
If those inputs are perfectly substitutable, 
the `new_k_ps()` function provides an answer. 

To demonstrate `new_k_ps()`, we'll create a very simple energy conversion chain.
(The `UKEnergyMatsWithR` ECC doesn't have the characteristic that we need, 
namely that an internediate industry has at least two reasonable-sized inputs.)
This ECC has two resource industries (R1 and R2),
one intermediate industry (I), and
two final demand sectors (Y1 and Y2).
The R1 industry makes product R1p.  
The R2 industry makes product R2p. 
Industry I makes product Ip.
Product Ip is consumed by final demand sectors Y1 and Y2.

The use (U), make (V), final demand (Y), and unit summation (S_units) matrices are given below.

```{r}
U <- matrix(c(0, 0, 10,
              0, 0, 10,
              0, 0,  0),
            byrow = TRUE, nrow = 3, ncol = 3,
            dimnames = list(c("R1p", "R2p", "Ip"), c("R1", "R2", "I"))) %>% 
  setrowtype("Products") %>% setcoltype("Industries")

V <- matrix(c(10,  0, 0,
               0, 10, 0, 
               0,  0, 4), 
            byrow = TRUE, nrow = 3, ncol = 3, 
            dimnames = list(c("R1", "R2", "I"), c("R1p", "R2p", "Ip"))) %>% 
  setrowtype("Industries") %>% setcoltype("Products")

Y <- matrix(c(0, 0,
              0, 0, 
              2, 2),
            byrow = TRUE, nrow = 3, ncol = 2,
            dimnames = list(c("R1p", "R2p", "Ip"), c("Y1", "Y2"))) %>% 
  setrowtype("Products") %>% setcoltype("Industries")

S_units <- matrix(c(1, 
                    1, 
                    1), 
                  byrow = TRUE, nrow = 3, ncol = 1,
                  dimnames = list(c("R1p", "R2p", "Ip"), c("quad"))) %>% 
  setrowtype("Products") %>% setcoltype("Units")
```

The Sankey diagram for this simple ECC is given below.

```{r, fig.width = 7, fig.height = 4}
make_sankey(U = U, V = V, Y = Y) %>% 
  extract2("Sankey")
```

We can change the proportion of inputs to industry I using `new_k_ps()`.
To do so, we must first calculate the input-output structure of the ECC 
using the `calc_io_mats()` function.
Then, we call the `new_k_ps()` function with a new k vector
that indicates industry I now uses 75/25 mixture of `R1` and `R2` instead of a 50/50 split between `R1` and `R2`.

```{r, fig.width = 7, fig.height = 4}
simple_iomats <- calc_io_mats(U = U, V = V, Y = Y, S_units = S_units)
k_prime <- matrix(c(0.75,
                    0.25), 
                  byrow = TRUE, nrow = 2, ncol = 1,
                  dimnames = list(c("R1p", "R2p"), c("I"))) %>% 
  setrowtype("Products") %>% setcoltype("Industries")
change_input_proportions <- new_k_ps(c(simple_iomats, list(U = U, V = V, Y = Y, S_units = S_units, 
                                                    k_prime = k_prime)))
make_sankey(U = change_input_proportions$U_prime, V = change_input_proportions$V_prime, Y = Y) %>% 
  extract2("Sankey")
```

If there were additional upstream industries, 
they would also change in response to the new ratio of inputs to industry `I`. 
The rate of resource extraction needed to supply `R1` would grow.
The rate of resource extraction needed to supply `R2` would shrink.


## Conclusion

`Recca` has three functions that allow an analyst to estimate the effect of changes to an ECC.

* `new_Y()` estimates changes upstream of final demand.
* `new_R_ps()` estimates changes downstream from resource extraction.
* `new_k_ps()` estimates changes upstream of inputs to an intermediate industry.

The first of these functions (`new_Y()`) was demonstrated in 
[Heun et al. [-@Heun:2018]](https://doi.org/10.1016/j.apenergy.2018.05.109).
The second and third functions (`new_R_ps()` and `new_k_ps()`) are shown here for the first time.


## References
