---
title: "new_* functions"
author: "Matthew Kuperus Heun"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{new_* functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(dplyr)
library(Recca)
library(matsbyname)
library(matsindf)
library(tidyr)
```

<!-- Establish some helpful LaTeX shortcuts for equations -->
\newcommand{\transpose}[1]{#1^\mathrm{T}}
\newcommand{\inverse}[1]{#1^{\mathrm{-}1}}
\newcommand{\mat}[1]{\mathbf{#1}}
\newcommand{\colvec}[1]{\mathbf{#1}}
\newcommand{\rowvec}[1]{\transpose{\colvec{#1}}}
\newcommand{\inversetranspose}[1]{\transpose{\left( \inverse{\mat{#1}} \right)}}
\newcommand{\transposeinverse}[1]{\inverse{\left( \transpose{\mat{#1}} \right)}}
\newcommand{\hatinv}[1]{\inverse{\widehat{#1}}}

## Introduction

`Recca` is an `R` package that builds upon input-output (I-O) analysis to analyze energy conversion chains (ECCs).
Historically, input-output (I-O) analysis has been used to analyze "what-if" scenarios involving
changes to the production structure of an economy. 
A common scenario asks "how will will the structure of intermediate industries
(represented by matrices $\mat{U}$ and $\mat{V}$) change when final demand (represented by matrix $\mat{Y}$) changes?"
Leontief's early work answered this question via a matrix inversion that now bears his name,
the "Leontief inverse ($\mat{L}$)." 
Analysts with an I-O background will know the famous equation $\mat{L} = (\mat{I} - \mat{A})^{-1}$.

The `Recca` package provides functions that perform several "what if" analyses for the analyst.
Each function has the name `new_*`, where `*` indicates the changed matrix or portion of the ECC.
Some functions have the name `new_*_ps` to indicate the functions assume perfect substitution (`ps`)
of industry inputs.
The `new_*` functions return the resource ($\mat{R}$), use ($\mat{U}$), make ($\mat{V}$), and final demand ($\mat{Y}$) matrices
that would be present under the conditions given by arguments to the `new_*` functions.

After a short discussion of the design philosophy of the `Reccca` package followed by  
some background information and setup, each type of analysis is described in turn.


## Design philosophy

The functions in `Recca` are flexible and can be used with either a data frame as input
(and names of columns as arguments to `Recca` function) or with matrices as inputs.
Outputs are either named columns appended to the input data frame or named matrices in a list.
The functions in `Recca` have default names for arguments which follow a standard nomenclature.
It is recommended that the default names be used whenever possible, 
thereby allowing cleaner code.


## Background and setup

For the examples that follow, we'll use the `UKEnergy2000mats` data frame.
Each row of `UKEnergy2000mats` data frame represents 
another version of a portion of the ECC for the UK in 2000.
Each `new_*` function requires basic knowledge about the I-O structure of the ECCs
in the form of named vectors and matrices.
The `calc_io_mats` function is the simplest way to obtain the I-O structure
which appears as additional columns appended at the right of the data frame.
The names of the additional columns is given by 
default argument values to the various `Recca` functions.

```{r}
library(dplyr)
library(matsbyname)
library(matsindf)
library(Recca)
library(tidyr)
IO_mats <- UKEnergy2000mats %>% 
  spread(key = "matrix.name", value = "matrix") %>% 
  # The following code is necessary until R matrices are included
  # in the UKEnergy2000mats data frame.
  rename(
    V_plus_R = V
  ) %>%
  separate_RV() %>% 
  # Obtain the I-O structure of the ECCs
  calc_io_mats()
# Note additional columns appended to the data frame.
glimpse(IO_mats)
```

Wtih the I-O structure of the ECC determined
(and represented by the $\colvec{q}$, $\colvec{f}$, and $\colvec{g}$ vectors and the 
$\mat{W}$, $\mat{Z}$, $\mat{K}$, $\mat{C}$, $\mat{D}$, $\mat{A}$, $\mat{L_{pxp}}$, and $\mat{L_{ixp}}$ matrices), 
we can use the `new_*` functions to ask "what if" questions.


## What if final demand changes? Function `new_Y()`

The classic I-O question is "what if final demand ($\mat{Y}$) changes?"


## What if perfectly-substitutable inputs to an intermediate industry change? Function `new_k_ps()`


## What if natural resource inputs to the ECC change? Function `new_R_ps()`


## Original text from `use_vignette()`

Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
