---
title: "Recca"
author: "Matthew Kuperus Heun"
date: "`r Sys.Date()`"
header-includes:
   - \usepackage{amsmath}
   - \usepackage{upgreek}
   - \usepackage{bm}
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Recca}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
bibliography: recca.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include = FALSE}
library(dplyr)
library(matsbyname)
library(matsindf)
library(Recca)
library(tibble)
library(tidyr)
```

<!-- Establish some helpful LaTeX shortcuts for equations -->

```{=tex}
\newcommand{\transpose}[1]{#1^\mathrm{T}}
\newcommand{\inverse}[1]{#1^{\mathrm{-}1}}
\newcommand{\mat}[1]{\mathbf{#1}}
\newcommand{\colvec}[1]{\mathbf{#1}}
\newcommand{\rowvec}[1]{\transpose{\colvec{#1}}}
\newcommand{\inversetranspose}[1]{\transpose{\left( \inverse{\mat{#1}} \right)}}
\newcommand{\transposeinverse}[1]{\inverse{\left( \transpose{\mat{#1}} \right)}}
\newcommand{\hatinv}[1]{\inverse{\widehat{#1}}}
```
```{=tex}
\newcommand{\matwithdims}[3]{\underset{\mathrm{#2} \times \mathrm{#3}}{\mat{#1}}}
\newcommand{\colvecwithdims}[2]{\underset{\mathrm{#2} \times 1}{\colvec{#1}}}
```
```{=tex}
\newcommand{\Lpp}{\matwithdims{L}{p}{p}}
\newcommand{\Lip}{\matwithdims{L}{i}{p}}
```


## Introduction

`Recca` (`R` Energy Conversion Chain Analysis) is an `R` package that
enables energy and exergy analysis of energy conversion chains. `Recca`
makes extensive use of a matrix-based Physical Supply Use Table (PSUT)
analysis technique that first appeared in [Heun et al.
[-@Heun:2018]](https://doi.org/10.1016/j.apenergy.2018.05.109). This
vignette walks through many of the calculations from [Heun et al.
[-@Heun:2018]](https://doi.org/10.1016/j.apenergy.2018.05.109), guided
by section numbers from the paper. We begin by discussing the design
philosophy of the `Recca` package, followed by a discussion of the PSUT
matrices that comprise an energy conversion chain (ECC). Thereafter,
calculation of the input-output structure of an ECC is demonstrated.
Finally, advanced calculations are shown, including changes in final
demand, net energy analysis, industry efficiencies, and energy
footprints.

## Design philosophy

The functions in `Recca` are flexibly designed and useful in many
situations. `Recca`'s flexibility stems from its extensive use of the
`matsbyname` and `matsindf` packages under the hood. Inputs to most
`Recca` functions can be any one of:

1.  matrices (with individual matrices as arguments to `Recca`
    functions),
2.  a list as the first argument (with names in the list indicating
    argument names), or
3.  a data frame (with names of data frame columns, as strings, as
    arguments to `Recca` functions).

Outputs are either named matrices in a list (for 1 and 2 above) or named
columns appended to the input data frame (for 3 above).

Argument names for matrices follow a standard nomenclature. It is
recommended that the default matrix names be used whenever possible,
thereby allowing cleaner code.

## PSUT matrices

(Reference [Heun et al.
[-@Heun:2018]](https://doi.org/10.1016/j.apenergy.2018.05.109), Section
2.2.2.)

For the examples that follow, we'll use the `UKEnergy2000tidy` data
frame. Each row of `UKEnergy2000tidy` data frame represents another
datum for a portion of the ECC for the UK in 2000. `UKEnergy2000tidy` is
in a format similar to data from the IEA or other organizations.

For data in the format of `UKEnergy2000tidy`, we can create `S_units`
matrices for each grouping using the `S_units_from_tidy()` function.

```{r}
library(tibble)
S_units <- UKEnergy2000tidy %>%
  dplyr::group_by(Country, Year, EnergyType, LastStage) %>%
  S_units_from_tidy()
tibble::glimpse(S_units)
```

And we can identify which entries belong in the resource ($\mathbf{R}$),
make ($\mathbf{V}$), use ($\mathbf{U}$), and final demand ($\mathbf{Y}$)
matrices with the `IEATools::add_psut_matnames()` and
`IEATools::add_row_col_meta()` functions.

```{r}
WithNames <- UKEnergy2000tidy %>%
  # Add a column indicating the matrix in which this entry belongs (U, V, or Y).
  IEATools::add_psut_matnames() %>%
  # Add metadata columns for row names, column names, row types, and column types.
  IEATools::add_row_col_meta() %>% 
  # Eliminate columns we no longer need
  dplyr::select(-LedgerSide, -FlowAggregationPoint, -Flow, -Product) %>%
  dplyr::mutate(
    # Ensure that all energy values are positive, as required for analysis.
    Edot = abs(Edot)
  )
head(WithNames)
```

After identifying the matrices, rownames, colnames, rowtypes, and
coltypes, we can collapse all data to matrices and add a unit summation
matrix (`S_units`).

```{r}
AsMats <- WithNames %>%
  # Collapse to matrices using functions in the matsindf package
  dplyr::group_by(Country, Year, EnergyType, LastStage, matnames) %>%
  matsindf::collapse_to_matrices(matnames = "matnames", matvals = "Edot",
                       rownames = "rownames", colnames = "colnames",
                       rowtypes = "rowtypes", coltypes = "coltypes") %>%
  dplyr::rename(matrix.name = matnames, matrix = Edot) %>%
  tidyr::spread(key = matrix.name, value = matrix) %>% 
  # Do a little more cleanup
  dplyr::mutate(
    # Create full U matrix
    U = matsbyname::sum_byname(U_feed, U_EIOU),
    # Create r_EIOU, a matrix that identifies the ratio of EIOU to other energy consumed.
    r_EIOU = matsbyname::quotient_byname(U_EIOU, U),
    r_EIOU = matsbyname::replaceNaN_byname(r_EIOU, val = 0)
  ) %>% 
  dplyr::select(-U_EIOU, -U_feed) %>%
  # Add S_units matrices
  dplyr::left_join(S_units, by = c("Country", "Year", "EnergyType", "LastStage")) %>%
  tidyr::gather(key = matrix.name, value = matrix, R, U, V, Y, r_EIOU, S_units)
tibble::glimpse(AsMats)
```

The `AsMats` data frame is essentially the same as the
`Recca::UKEnergy2000mats` data frame. The remainder of this vignette
uses the `UKEnergy2000mats` data frame.

## I-O structure

(Reference [Heun et al.
[-@Heun:2018]](https://doi.org/10.1016/j.apenergy.2018.05.109), Section
2.2.4.)

### With individual matrices

To determine the I-O structure of an ECC, use the `calc_io_mats()`
function.

```{r}
library(tidyr)
mats <- UKEnergy2000mats %>% 
  tidyr::spread(key = matrix.name, value = matrix) %>% 
  # Put rows in a natural order
  dplyr::mutate(
    LastStage = factor(LastStage, levels = c("Final", "Useful", "Services")),
    EnergyType = factor(EnergyType, levels = c("E", "X"))
  ) %>% 
  dplyr::arrange(LastStage, EnergyType)
# Use the calc_io_mats function with individual matrices, 
# each taken from the first row of the UKEnergy2000mats data frame.
R <- mats$R[[1]]
U <- mats$U[[1]]
U_feed = mats$U_feed[[1]]
V <- mats$V[[1]]
Y <- mats$Y[[1]]
S_units <- mats$S_units[[1]]
IO_list <- calc_io_mats(R = R, U = U, U_feed = U_feed, V = V, Y = Y, S_units = S_units)
```

Most `Recca` functions return a list when called with individual
matrices as arguments. The `calc_io_mats()` function gives several I-O
matrices in its returned list.

```{r}
class(IO_list)
names(IO_list)
IO_list[["y"]]
```

The same calculations can be performed by supplying a named list to
`Recca` functions. In this approach, all original data are returned in
the list. So in this case, matrices $\mathbf{U}$, $\mathbf{V}$,
$\mathbf{Y}$, and $\mathbf{S_{units}}$ are also returned from the
`calc_io_mats()` function. When a list is supplied to a `Recca` function
in the `.sutmats` argument, most other input arguments must be strings
that identify the names of appropriate entries in the `.sutmats` list
containing named vectors or matrices. Helpfully, the default values for
other input arguments conform to a standard nomenclature. When using the
standard nomenclature, most `Recca` functions can use the default
arguments for input and output items in the list.

```{r}
IO_from_list <- calc_io_mats(list(R = R, U = U, U_feed = U_feed, V = V, Y = Y, S_units = S_units))
class(IO_from_list)
names(IO_from_list)
IO_from_list[["y"]]
```

### From a `matsindf`-style data frame

Most `Recca` functions can also operate on a `matsindf`-style data
frame. (A `matsindf`-style data frame has matrices in cells of a data
frame. See the `matsindf` package for additional information.) When a
data frame is supplied to a `Recca` function in the `.sutmats` argument,
most other input arguments must be strings that identify the names of
appropriate columns in `.sutmats` containing named vectors or matrices.
Helpfully, the default values for other input arguments conform to a
standard nomenclature. When using the standard nomenclature, most
`Recca` functions can use the default arguments for input and output
columns. This approach yields very clean piped code, as shown below.

To illustrate the above features of `Recca` functions, we'll apply the
`calc_io_mats` function to the entire `UKEnergy2000mats` data frame,
calculating appropriate I-O matrices for each row. Used in this way,
`Recca` functions act like specialized `dplyr::mutate()` functions, with
new columns added to the right side of the data frame supplied to the
`.sutmats` argument.

```{r}
IO_df <- mats %>% calc_io_mats()
```

By inspecting `IO_df`, we can see, for example, that one $\mathbf{y}$
vector was calculated for each of the four rows of `mats`. The same is
true for all I-O matrices calculated by `calc_io_mats`.

```{r}
class(IO_df)
names(IO_df)
glimpse(IO_df)
IO_df[["y"]][[1]]
IO_df[["y"]][[4]]
```

For the remainder of this vignette, operations will be performed on the
entire `UKEnergy2000mats` data frame. But readers should remember that
functions can be called on named lists or individual matrices, as well.

## Changes in final demand

(Reference [Heun et al.
[-@Heun:2018]](https://doi.org/10.1016/j.apenergy.2018.05.109), Section
2.2.5.)

One of the first applications of input-output analysis was estimating
changes in industry outputs that would be required to meet new final
demand. `Recca` allows similar calculations on energy conversion chains
with the function `new_Y()`. Arguments to `new_Y()` include matrices
that describe the input-output structure of the ECC and `Y_prime`, the
new final demand matrix. `new_Y()` calculates `U_prime` and `V_prime`
matrices which represent the ECC that would be required to meet the new
final demand represented by `Y_prime`.

```{r}
Double_demand <- IO_df %>% 
  dplyr::mutate(
    Y_prime = matsbyname::hadamardproduct_byname(2, Y)
  ) %>% 
  new_Y()
names(Double_demand)
IO_df[["Y"]][[1]][ , c(1,2)]
Double_demand[["Y_prime"]][[1]]
IO_df[["U"]][[1]][ , c("Crude dist.", "Diesel dist.")]
Double_demand[["U_prime"]][[1]][ , c("Crude dist.", "Diesel dist.")]
```

See the vignette for the [`new_*()` functions](new-functions.html) for
more details.

## Net energy analysis

(Reference [Heun et al.
[-@Heun:2018]](https://doi.org/10.1016/j.apenergy.2018.05.109), Section
3.1.)

The energy production system itself consumes energy, and important
metrics for energy conversion chain industries are energy return ratios
(ERRs). `Recca` provides a function to calculate three ERRs, a gross
energy ratio (GER), a net energy ratio (NER), and the ratio of NER to
GER. GER is commonly called energy return on investment (EROI). (See
[Brandt & Dale [-@Brandt:2011]](https://doi.org/10.3390/en4081211).)
These ERRs can be calculated for a variety of system boundaries. ERRs
for the $\gamma$ system boundary can be calculated readily using the
`calc_ERRs_gamma()` function. All three ERRs are calculated at the same
time. The ERRs are `NA` for industries in which inputs or outputs are
unit-inhomogeneous. The ERRs are `Inf` for industries that make an
energy product without consuming any energy from another processing
chain (such as the Elect. grid). The results below are identical to Fig.
6 in [Heun et al.
[-@Heun:2018]](https://doi.org/10.1016/j.apenergy.2018.05.109).

```{r}
ERRs <- IO_df %>% 
  calc_ERRs_gamma()
ERRs$ger_gamma[[1]]
ERRs$ner_gamma[[1]]
ERRs$r_gamma[[1]]
```

## Efficiencies

(Reference [Heun et al.
[-@Heun:2018]](https://doi.org/10.1016/j.apenergy.2018.05.109), Section
3.2.)

The efficiency of every industry in the ECC can be calculated quickly
with the `calc_eta_i()` function, which creates a column named `eta_i`
(by default) at the right of the data frame. If a particular ECC has
industries whose inputs or outputs are unit inhomogeneous, the `eta_i`
vector will have `NA` values in the appropriate places. The results
below are identical to Fig. 9 in [Heun et al.
[-@Heun:2018]](https://doi.org/10.1016/j.apenergy.2018.05.109).

```{r}
etas <- IO_df %>% 
  calc_eta_i()
names(etas)
etas[["eta_i"]][[1]]
etas[["eta_i"]][[3]] # NAs indicate inhomogeneous units on inputs or outputs.
```

## Energy footprints

(Reference [Heun et al.
[-@Heun:2018]](https://doi.org/10.1016/j.apenergy.2018.05.109), Section
3.3.)

Final demand for energy contains embodied primary energy, the sum of all
primary energy consumed and wasted throughout the ECC in the process of
satisfying that final demand. `Recca` provides two functions to
calculate embodied primary energy and important ratios, namely
`calc_embodied_mats()` and `calc_embodied_etas()`.

The function `calc_embodied_mats()` calculates embodied energy in final
demand products ($p$) and final demand sectors ($s$).

Outputs from `calc_embodied_mats()` include the following matrices:

+--------------------------------------+-----------------------------------------------------------------+
| Matrix                               | Description (rows$\times$columns)                               |
+=====================================:+:================================================================+
| $\mathbf{G} =                        | Industry and resource stocks output requirements for final      |
| \mathbf{G}_R + \mathbf{G}_V$         | demand ($(r + i) \times p$)                                     |
+--------------------------------------+-----------------------------------------------------------------+
| $\mathbf{G}_R =                      | Resource stocks output requirements for final demand            |
| \mathbf{R}                           | ($r \times p$)                                                  |
| \widehat{\mathbf{q}}^{\mathrm{-}1}   |                                                                 |
| \underset{\mathrm{p} \times          |                                                                 |
| \mathrm{p}}{\mathbf{L}}              |                                                                 |
| \hat{\mathbf{y}}$                    |                                                                 |
+--------------------------------------+-----------------------------------------------------------------+
| $\mathbf{G}_V =                      | Industry output requirements for final demand ($i \times p$)    |
| \underset{\mathrm{i} \times          |                                                                 |
| \mathrm{p}}{\mathbf{L}}              |                                                                 |
| \widehat{\mathbf{y}}$                |                                                                 |
+--------------------------------------+-----------------------------------------------------------------+
| $\mathbf{H} =                        | Industry and resource stocks output requirements for final      |
| \mathbf{H}_R + \mathbf{H}_V$         | demand sectors ($(r + i) \times s$)                             |
+--------------------------------------+-----------------------------------------------------------------+
| $\mathbf{H}_R =                      | Resource stocks output requirements for final demand sectors    |
| \mathbf{R}                           | ($r \times s$)                                                  |
| \widehat{\mathbf{q}}^{\mathrm{-}1}   |                                                                 |
| \underset{\mathrm{p} \times          |                                                                 |
| \mathrm{p}}{\mathbf{L}}\mathbf{Y}$   |                                                                 |
+--------------------------------------+-----------------------------------------------------------------+
| $\mathbf{H}_V =                      | Industry output requirements for final demand sectors           |
| \underset{\mathrm{i} \times          | ($i \times s$)                                                  |
| \mathrm{p}}{\mathbf{L}}\mathbf{Y}$   |                                                                 |
+--------------------------------------+-----------------------------------------------------------------+
| $\mathbf{E} =                        | Energy or services produced ($+$) or consumed ($-$) per unit    |
| [ (\mathbf{R} +                      | output by industries (in columns) ($p \times i$)                |
| \mathbf{V})^\mathrm{T} -             |                                                                 |
| \mathbf{U}_{feed} ]                  |                                                                 |
| \widehat{(\mathbf{r} +               |                                                                 |
| \mathbf{g})}^{-1}$                   |                                                                 |
+--------------------------------------+-----------------------------------------------------------------+
| $\mathbf{r} =                        | Row sums of $\mathbf{R}$ matrix; works only when $\mathbf{R}$   |
| \bar{\mathbf{R}} \mathbf{i}$         | is unit-homogeneous                                             | 
+--------------------------------------+-----------------------------------------------------------------+
| $\mathbf{g} =                        | Row sums of $\mathbf{V}$ matrix; works only when $\mathbf{V}$   |
| \bar{\mathbf{V}} \mathbf{i}$         | is unit-homogeneous                                             |
+--------------------------------------+-----------------------------------------------------------------+
| $\mathbf{e}_i$                       | Rows of $\mathbf{E}$; subscript $i$ indicates energy products   |
+--------------------------------------+-----------------------------------------------------------------+
| $\mathbf{Q}_i =                      | Sources (positive entries) and                                  |
| \widehat{\mathbf{e}_i} \mathbf{G}$   | consumption (negative entries) by industries (in rows)          |
|                                      | of embodied energy products (subscript $i$) in consumed         |
|                                      | energy products (in columns)                                    |
+--------------------------------------+-----------------------------------------------------------------+
| $\mathbf{Q}_i^+$                     | $\mathbf{Q}_i$ with negative entries set to zero                |
+--------------------------------------+-----------------------------------------------------------------+
| $\mathbf{i} \mathbf{Q}_i^+$          | Column sums (to form row vectors) of $\mathbf{Q}_i^+$           |
|                                      | (embodied product $\times$ embodying products                   |
+--------------------------------------+-----------------------------------------------------------------+
| $\mathbf{M}_p$                       | Each row is one $\mathbf{i} \mathbf{Q}_i^+$ to                  |
|                                      | show embodied energy products in each embodying energy product  |
|                                      | (embodied products $\times$ embodying products)                 |
+--------------------------------------+-----------------------------------------------------------------+
| $\mathbf{M}_s = \mathbf{M}_p         | Embodied energy products consumed by final demand sectors       |
| \hat{\mathbf{q}}^{-1} \mathbf{Y}$    | (embodied products $\times$ consuming final demand sectors)     |
+--------------------------------------+-----------------------------------------------------------------+

`calc_embodied_mats()` also calculates $\mathbf{F}_{footprint}$ and
$\mathbf{F}_{ef\!fects}$ matrices, which answer the questions
(respectively) "What is the fractional composition of embodied energy of
each final demand energy type?" and "What is the fractional destination
of a given upstream energy product?" The calculations can be performed
for each final demand product or each final demand sector. The following
table describes these matrices.

+--------------------------------------------+-----------------------------------------------------------------+
| Matrix                                     | Description                                                     |
+===========================================:+:================================================================+
| $\mathbf{F}_{footprint,p} =                | Each final demand product (columns) contains embodied energy.   |
| \mathbf{M}_p                               | On a fractional basis, where does that embodied energy come     |
| (\widehat{\mathbf{i} \mathbf{M}_p})^{-1}$  | from (rows)? Columns sum to 1.                                  |
+--------------------------------------------+-----------------------------------------------------------------+
| $\mathbf{F}_{ef\!fects,p} =                | Each upstream energy product (rows) becomes embodied somewhere. |
| (\widehat{\mathbf{M}_p} \mathbf{i})^{-1}   | On a fractional basis, where does that upstream energy become   |
| \mathbf{M}_p$                              | embodied (columns)? Rows sum to 1.                              |
+--------------------------------------------+-----------------------------------------------------------------+
| $\mathbf{F}_{footprint,s} =                | Each final demand sector (columns) consumes embodied energy. On |
| \mathbf{M}_s                               | a fractional basis, where does the consumed embodied energy     |
| (\widehat{\mathbf{i} \mathbf{M}_s})^{-1}$  | come from (rows)? Columns sum to 1.                             |
+--------------------------------------------+-----------------------------------------------------------------+
| $\mathbf{F}_{ef\!fects,s} =                | Each embodied upstream energy product (rows) is consumed by a   |
| (\widehat{\mathbf{M}_s} \mathbf{i})^{-1}   | final demand sector. On a fractional basis, where is that       |
| \mathbf{M}_s$                              | embodied upstream energy consumed (columns)? Rows sum to 1.     |
+--------------------------------------------+-----------------------------------------------------------------+

Finally, embodied energy efficiencies can be calculated as final demand
energy divided by embodied energy. Again, efficiencies can be calculated
for each final demand product or each final demand sector. The
`calc_embodied_etas()` function does this computation.

+------------------------------------------------------+-----------------------------------------------------------------+
| Matrix                                               | Description                                                     |
+=====================================================:+:================================================================+
| $\eta_p = (\widehat{\mathbf{G}^\mathrm{T}            | Final demand product (rows) divided by embodied primary         |
| \mathbf{s}_r})^{-1} \mathbf{y}$                      | energy of that final demand product                             |
+------------------------------------------------------+-----------------------------------------------------------------+
| $\eta_s = (\widehat{\mathbf{s}_r \mathbf{H}})^{-1}   | Energy consumed by final demand sector (rows) divided by        |
| (\mathbf{iY})^{\mathrm{T}}$                          | embodied energy input to the final demand sector                |
+------------------------------------------------------+-----------------------------------------------------------------+

Note: $\mathbf{s}_r$ is a selection vector for resource industries.

```{r}
primary_machine_names <- c("Resources - Crude", "Resources - NG")

embodied_mats <- IO_df %>% 
  dplyr::mutate(
    U_EIOU = matsbyname::hadamardproduct_byname(r_EIOU, U)
  ) %>%
  calc_embodied_mats() %>%
  calc_embodied_etas(primary_machine_names = primary_machine_names)
names(embodied_mats)
```

Figure 15 in [Heun et al.
[-@Heun:2018]](https://doi.org/10.1016/j.apenergy.2018.05.109) shows
primary-to-services energetic efficiencies for the ECC in the 3rd row of
`IO_df`. The following code extracts those results. Rows of the vector
give final demand services and their units. The column gives
efficiencies in units of service per ktoe of energy.

```{r}
embodied_mats$eta_p[[3]]
```

## Conclusion

This vignette demonstrated the use of the `Recca` package. `Recca`
provides many useful functions for analyzing energy conversion chains
within the PSUT framework first described in [Heun et al.
[-@Heun:2018]](https://doi.org/10.1016/j.apenergy.2018.05.109).

## References
